Using an iterated block particle filter via \pkg{spatPomp}

### 2024-08-06 ms.Rnw
edits based on Lees review - adding purled script

%% basic POMP definitions
\newcommand\Xspace{{\mathbb X}}
\newcommand\Yspace{{\mathbb Y}}
%\newcommand\Thetaspace{{\Theta}}
%\newcommand\vecTheta{\vec{\Theta}}
\newcommand\vecTheta{\Theta}
\newcommand\Thetaspace{\R^{\Thetadim}}
\newcommand\hatTheta{\widehat{\Theta}}
\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
\renewcommand\Xdim{D_X}
\renewcommand\Xspace{{\R^{\Xdim}}}
\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
\renewcommand\Yspace{\R}
%\newcommand\Thetadim{{\mathrm{dim}}(\Thetaspace)}
\newcommand\Thetadim{D_\theta}
\newcommand\thetadim{d_\theta}
%\newcommand\vectheta{\vec{\theta}}
\newcommand\vectheta{\theta}
\newcommand\timeSet{{\mathbb T}}
\newcommand\data[1]{#1^*}

@@

%% for bagged filters
\newcommand\ABF{ABF}
\newcommand\ABFIR{ABF-IR}
\newcommand\UBF{UBF}
\newcommand\rep{i}
\newcommand\altRep{\tilde\rep}
\newcommand\Rep{\mathcal{I}}

@@

%% customized math macros
\newcommand\prob{\mathbb{P}}
\newcommand\dd[1]{\mathrm{d}{#1}}
\newcommand\given{{\,\vert\,}}
\newcommand\equals{{{\,}={\,}}}
\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
\newcommand\myto{{\;:\;}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\cp[2]{N_{\mathrm{#1}\mathrm{#2}}}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\argequals{{\,=\,}}
\newcommand\lags{c}
\newcommand\maxlag{\overline{c}}
\newcommand\nlfList{C}
\newcommand\bigO{\mathcal{O}}
\newcommand\loglik{\lambda}
\newcommand\loglikMC{\MC{\loglik}}
\newcommand\loglikComponent{c}
\newcommand\R{\mathbb{R}}
\newcommand\param{\,;}
\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
\newcommand\MC[1]{#1^{\mbox{\tiny MC}}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{\mathrm{dist}}
\newcommand\transpose{t}

@@

<<packages,include=F,echo=F,cache=F>>=
library("spatPomp")
library("ggplot2")
library("tidyverse")
library("knitr")
stopifnot(packageVersion("pomp")>="4.1")
@

@@

<<knitr-opts,include=F,cache=F,purl=F>>=
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
#    tidy=TRUE,
#    tidy.opts=list(
#        keep.blank.line=FALSE
#    ),
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
#    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
    fig.height=5,fig.width=8,
    out.width="\\textwidth",
    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=12
    )
)
@

@@

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\small This version compiled on \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
Source code at \url{https://github.com/ionides/ibpf-tutorial}}
\begin{document}
\maketitle

\begin{abstract}
The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
In addition to the \pkg{spatPomp} code presented here, the full code to reproduce this document is available in the R Noweb (\code{.Rnw}) source file. 
\end{abstract}

@@

<<bm-followup,echo=F,eval=T>>=
if(i==1) cores <- 2   # no need to use all available cores when i=1
U <- length(unit_names(b))
@

Here, \code{i} is a computational intensity switch which adjusts the code for varying run-time.
We set \code{i=1} for testing and debugging, and \code{i=2} for higher quality results.
For simplicity, we consider only one unit-specific parameter, $\rho_{\unit}$, with other parameters being fixed at a value shared between units.
The simulation for \code{b} has $\rho_u=\Sexpr{coef(b)["rho1"]}$ for all $\unit$, but the estimators do not know this.
Before carrying out inference, we check likelihood evaluation.
For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
This study uses a sufficiently small number of units (\code{U=\Sexpr{U}}) that the particle filter is numerically tractable.
We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
We can readily validate the agreement between \code{bm2\_kalman\_logLik} and \code{pfilter}, and identify the likelihood cost of the block filter approximation in this situation.

<<bm_logLik_code,cache=F,echo=T,eval=F>>=
kf_logLik <- bm2_kalman_logLik(b)
pf_logLik <- replicate(10,
  logLik(pfilter(b,switch(i,10,1000)))
)
bpf_logLik2 <- replicate(10,
  logLik(bpfilter(b,switch(i,10,1000),block_size=2))
)
@

@@

\citet{ionides21} investigated a range of values for $U$ for this model in their Figure~1, and for an epidemiological model their Figure~3.
The small scenario considered here, with $U=4$, is designed for the following purposes: (i) to validate whether or not \code{ibpf} is correctly coded by comparison with direct calculations using the Kalman filter; (ii) to check whether or not the block approximation has considerable adverse effects on inference in this case.
The inherent scalability of BPF and IBPF, together with the spatial homogeneity of the model, suggests that results for $U=4$ should be representative for larger $U$.

@@

prof_init <- rep(0.4,U)
names(prof_init) <- paste0("rho",1:U)
prof_vals <- seq(from=0,to=0.8,length=switch(i,5,20))

@@

<<ibpf-mle-code,eval=F,echo=T>>=
rho_start <- seq(from=0.2,to=0.8,length=U)
params_start <- coef(b)
params_start[paste0("rho",1:U)] <- rho_start
ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
  ibpf(b,params=params_start,
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=rw_sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
@

@@

<<ibpf-prof,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof1.rda"),seed=722,{
  tic <- Sys.time()
  u <- 1
  rho_start <- seq(from=0.2,to=0.8,length=U)
  reps <- switch(i,3,10)
  param_start_matrix <- matrix(params_start,byrow=TRUE,
    ncol=length(coef(b)),nrow=reps*length(prof_vals),
    dimnames=list(NULL,names(coef(b))))
  param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
  ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
    rw_sd_call <- rep(list(0.02),times=U-1)
    names(rw_sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
    ibpf(b,
      params=param_start_matrix[s,],
      Nbpf=switch(i,2,50),Np=switch(i,20,1000),
      rw.sd=do.call(rw_sd,rw_sd_call),
      unitParNames="rho",
      sharedParNames=NULL,
      block_size=2,
      cooling.fraction.50=0.5
    )
  }
  toc <- Sys.time()
})
prof1time <- toc-tic
@

@@

prof_sig_init <- rep(1,U)
names(prof_sig_init) <- paste0("sigma",1:U)
prof_sig_vals <- seq(from=0.4,to=1.6,length=switch(i,5,20))

@@

<<ibpf-prof-sig,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof_sig1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
sigma_start <- seq(from=0.4,to=1.6,length=U)
params_sig_start <- coef(b_sig)
params_sig_start[paste0("sigma",1:U)] <- sigma_start
reps <- switch(i,2,10)
param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
  dimnames=list(NULL,names(coef(b_sig))))
param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
  rw_sd_call <- rep(list(0.02),times=U-1)
  names(rw_sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
  ibpf(b_sig,
    params=param_sig_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=do.call(rw_sd,rw_sd_call),
    unitParNames="sigma",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof_sig1time <- toc-tic
@

@@

<<model-he10-code,eval=F,echo=T>>=
he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
  expandedParNames=c("R0"),
  towns_selected=c(1,2,11,12),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=800,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m <- simulate(he10_model,seed=27)
@

@@

<<m-prof-R0,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_prof_R02.rda"),seed=722,{
m_prof_R0_vals <- seq(from=25,to=35,length=switch(i,5,20))
tic <- Sys.time()
m_R0 <- m
m_R0_prof_u <- 1
reps <- switch(i,3,10)
param_R0_start_matrix <- matrix(coef(m_R0),byrow=TRUE,
  ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
  dimnames=list(NULL,names(coef(m_R0))))
param_R0_start_matrix[,paste0("R0",m_R0_prof_u)] <- rep(m_prof_R0_vals,each=reps)
m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
  rw_sd_call <- rep(list(0.02),times=U-1)
  names(rw_sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
  ibpf(m_R0,
    params=param_R0_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
    rw.sd=do.call(rw_sd,rw_sd_call),
    unitParNames="R0",
    sharedParNames=NULL,
    block_size=1,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
prof_R0_cores <- cores
})
m_prof_R0_time <- toc-tic
@

@@

<<m-slice-g,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_slice_g.rda"),seed=722,{
m_g_prof_u <- 4
slice_reps <- switch(i,2,10)
slice_points <- switch(i,5,20)
slice_g_vals <- rep(seq(from=100,to=2500,length=slice_points),each=slice_reps)
params_g_slice<- matrix(coef(m_g),
  nrow=slice_reps*slice_points,
  ncol=length(coef(m_g)),
  byrow=T,
  dimnames=list(NULL,names(coef(m_g))))
params_g_slice[,paste0("g",m_g_prof_u)] <- slice_g_vals
tic <- Sys.time()
m_slice_g_evals <- foreach(s=1:nrow(params_g_slice),.combine=c)%dopar%{
  logLik(bpfilter(m_g,
    params=params_g_slice[s,],
    Np=switch(i,10,5000),
    block_size=1
  ))
}
toc <- Sys.time()
m_slice_g_time <- toc-tic
m_slice_cores <- cores
})
@

@@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observation.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citet{asfaw23arxiv}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@

>>>>

%% basic POMP definitions
\newcommand\Xspace{{\mathbb X}}
\newcommand\Yspace{{\mathbb Y}}
%\newcommand\Thetaspace{{\Theta}}
%\newcommand\vecTheta{\vec{\Theta}}
\newcommand\vecTheta{\Theta}
\newcommand\Thetaspace{\R^{\Thetadim}}
\newcommand\hatTheta{\widehat{\Theta}}
\newcommand\Xdim{{\mathrm{dim}}(\Xspace)}
\renewcommand\Xdim{D_X}
\renewcommand\Xspace{{\R^{\Xdim}}}
\newcommand\Ydim{{\mathrm{dim}}(\Yspace)}
\renewcommand\Yspace{\R}
\newcommand\Thetadim{D_\theta}
\newcommand\thetadim{d_\theta}
\newcommand\vectheta{\theta}
\newcommand\timeSet{{\mathbb T}}
\newcommand\data[1]{#1^*}

@@

@@

%% customized math macros
\newcommand\prob{\mathbb{P}}
\newcommand\dd[1]{\mathrm{d}{#1}}
\newcommand\given{{\,\vert\,}}
\newcommand\equals{{{\,}={\,}}}
\newcommand\myequals{\hspace{0.5mm}{=}\hspace{0.5mm}}
\newcommand\myto{{\;:\;}}
\newcommand\seq[2]{{#1}\!:\!{#2}}
\newcommand\mydot{{\,\cdot\,}}
\newcommand\giventh{{\hspace{0.5mm};\hspace{0.5mm}}}
\newcommand\normal{{\mathrm{Normal}}}
\newcommand\argequals{{\,=\,}}
\newcommand\bigO{\mathcal{O}}
\newcommand\loglik{\lambda}
\newcommand\loglikMC{\MC{\loglik}}
\newcommand\loglikComponent{c}
\newcommand\R{\mathbb{R}}
\newcommand\param{\,;}
\newcommand\mycolon{{\hspace{0.6mm}:\hspace{0.6mm}}}
\newcommand\MC[1]{#1^{\mbox{\tiny MC}}}
\newcommand\Var{\mathrm{Var}}
\newcommand\var{\Var}
\newcommand\Cov{\mathrm{Cov}}
\newcommand\cov{\Cov}
\newcommand\iid{\mathrm{iid}}
\newcommand\dist{\mathrm{dist}}
\newcommand\transpose{t}

@@

<<packages,include=F,echo=F,cache=F>>=
library("spatPomp")
library("ggplot2")
library("tidyverse")
library("knitr")
stopifnot(packageVersion("pomp")>="5.0")
@

@@

<<knitr-opts,include=F,cache=F,purl=F>>=
opts_chunk$set(
    progress=TRUE,prompt=TRUE,highlight=FALSE,
    comment="",
    warning=FALSE,
    message=FALSE,
    error=TRUE,
    echo=TRUE,
    cache=FALSE,  ### SET CACHE TO TRUE WHILE EDITING TEXT
#    cache=TRUE,  ### SET CACHE TO FALSE WHILE DEBUGGING CODE
    strip.white=TRUE,
    results="markup",
    background="#FFFFFF00",
    size="normalsize",
    fig.path="figure/",
    fig.lp="fig:",
    fig.align="left",
    fig.show="asis",
    fig.height=5,fig.width=8,
    out.width="\\textwidth",
    dpi=300,
    dev="pdf",
    dev.args=list(
        bg="transparent",
        pointsize=12
    )
)
@

@@

\title{Using an iterated block particle filter via \pkg{spatPomp}}
\author{Ning Ning and Edward L. Ionides}
\date{\small This version compiled on \today,  using \proglang{R} \Sexpr{paste0(R.version$major,".",R.version$minor)}, \pkg{spatPomp} \Sexpr{packageVersion('spatPomp')}, and \pkg{pomp} \Sexpr{packageVersion('pomp')}\\
\vspace{3mm}
Source code at \url{https://github.com/ionides/ibpf-tutorial}\\
\vspace{3mm}
R script at \url{https://github.com/ionides/ibpf-tutorial/blob/main/ms.R}}

\begin{document}
\maketitle

\begin{abstract}
The IBPF algorithm studied by \citet{ning23-ibpf} and \citet{ionides22} has been contributed to the \proglang{R} package \pkg{spatPomp} \citep{asfaw21cran,asfaw23arxiv} as the function  \code{ibpf}.
This document introduces \code{ibpf} and validates its correctness on a simple Gaussian example which is tractable using the Kalman filter.
We also test \code{ibpf} on simulated data for a measles transmission model.
\end{abstract}

@@

<<bm-followup,echo=F,eval=T>>=
if(i==1) cores <- 2   # no need to use all available cores when i=1
@

Here, \code{i} is a computational intensity switch which adjusts the code for varying run-time.
We set \code{i=1} for testing and debugging, and \code{i=2} for higher quality results.
For simplicity, we consider only one unit-specific parameter, $\rho_{\unit}$, with other parameters being fixed at a value shared between units.
The simulation for \code{b} has $\rho_u=\Sexpr{coef(b)["rho1"]}$ for all $\unit$, but the estimators do not know this.
Before carrying out inference, we check likelihood evaluation.
For this toy model, the \pkg{spatPomp} function \code{bm2\_kalman\_logLik} provides an exact log-likelihood via the Kalman filter.
This study uses a small number of units,
<<define-U,echo=T>>=
U <- 4
@
so that the particle filter is numerically tractable.
We use the particle filter provided by the \pkg{pomp} package \citep{king16}, taking advantage of the class structure where \class{spatPomp} inherits from \class{pomp}.
We can readily validate the agreement between \code{bm2\_kalman\_logLik} and \code{pfilter}, and identify the likelihood cost of the block filter approximation in this situation.

<<bm_logLik_code,cache=F,echo=T,eval=F,purl=F>>=
kf_logLik <- bm2_kalman_logLik(b)
pf_logLik <- replicate(10,
  logLik(pfilter(b,switch(i,10,1000)))
)
bpf_logLik2 <- replicate(10,
  logLik(bpfilter(b,switch(i,10,1000),block_size=2))
)
@

@@

\citet{ionides21} investigated a range of values for $U$ for this model in their Figure~1, and for an epidemiological model their Figure~3.
The small scenario considered here, with $U=\Sexpr{U}$, is designed for the following purposes: (i) to validate whether or not \code{ibpf} is correctly coded by comparison with direct calculations using the Kalman filter; (ii) to check whether or not the block approximation has considerable adverse effects on inference in this case.
The inherent scalability of BPF and IBPF, together with the spatial homogeneity of the model, suggests that results for $U=\Sexpr{U}$ should be representative for larger $U$.

@@

prof_init <- rep(0.4,U)
names(prof_init) <- paste0("rho",1:U)
prof_vals <- seq(from=0,to=0.8,length.out=switch(i,5,20))

@@

<<ibpf-mle-code,eval=F,echo=T,purl=F>>=
rho_start <- seq(from=0.2,to=0.8,length.out=U)
params_start <- coef(b)
params_start[paste0("rho",1:U)] <- rho_start
ibpf_mle_searches <- foreach(reps=1:switch(i,3,10))%dopar%{
  ibpf(b,params=params_start,
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=rw_sd(rho1=0.02,rho2=0.02,rho3=0.02,rho4=0.02),
    unitParNames="rho",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
@

@@

<<ibpf-prof,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof1.rda"),seed=722,{
  tic <- Sys.time()
  u <- 1
  rho_start <- seq(from=0.2,to=0.8,length.out=U)
  reps <- switch(i,3,10)
  param_start_matrix <- matrix(params_start,byrow=TRUE,
    ncol=length(coef(b)),nrow=reps*length(prof_vals),
    dimnames=list(NULL,names(coef(b))))
  param_start_matrix[,paste0("rho",u)] <- rep(prof_vals,each=reps)
  ibpf_prof_searches <- foreach(s=1:nrow(param_start_matrix))%dopar%{
    rw_sd_call <- rep(list(0.02),times=U-1)
    names(rw_sd_call) <- lapply((1:U)[-u],function(x)paste0("rho",x))
    ibpf(b,
      params=param_start_matrix[s,],
      Nbpf=switch(i,2,50),Np=switch(i,20,1000),
      rw.sd=do.call(rw_sd,rw_sd_call),
      unitParNames="rho",
      sharedParNames=NULL,
      block_size=2,
      cooling.fraction.50=0.5
    )
  }
  toc <- Sys.time()
})
prof1time <- toc-tic
@

@@

prof_sig_init <- rep(1,U)
names(prof_sig_init) <- paste0("sigma",1:U)
prof_sig_vals <- seq(from=0.4,to=1.6,length.out=switch(i,5,20))

@@

<<ibpf-prof-sig,eval=T,echo=F>>=
stew(file=paste0(bm_dir,"prof_sig1.rda"),seed=722,{
tic <- Sys.time()
u <- 1
sigma_start <- seq(from=0.4,to=1.6,length.out=U)
params_sig_start <- coef(b_sig)
params_sig_start[paste0("sigma",1:U)] <- sigma_start
reps <- switch(i,2,10)
param_sig_start_matrix <- matrix(params_sig_start,byrow=TRUE,
  ncol=length(coef(b_sig)),nrow=reps*length(prof_sig_vals),
  dimnames=list(NULL,names(coef(b_sig))))
param_sig_start_matrix[,paste0("sigma",u)] <- rep(prof_sig_vals,each=reps)
ibpf_prof_sig_searches <- foreach(s=1:nrow(param_sig_start_matrix))%dopar%{
  rw_sd_call <- rep(list(0.02),times=U-1)
  names(rw_sd_call) <- lapply((1:U)[-u],function(x)paste0("sigma",x))
  ibpf(b_sig,
    params=param_sig_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,1000),
    rw.sd=do.call(rw_sd,rw_sd_call),
    unitParNames="sigma",
    sharedParNames=NULL,
    block_size=2,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
})
prof_sig1time <- toc-tic
@

@@

<<model-he10-code,eval=F,echo=T,purl=F>>=
he10_model <- he10(U=4,dt=1/365,Tmax=switch(i,1955,1964),
  expandedParNames=c("R0"),
  towns_selected=c(1,2,11,12),
  basic_params = c(
    alpha =0.99,      iota=0,          R0=30,
    cohort=0.5,  amplitude=0.3,     gamma=52,
    sigma=52,           mu=0.02,  sigmaSE=0.05,
    rho=0.5,           psi=0.1,         g=800,
    S_0=0.036,         E_0=0.00007,   I_0=0.00006
  )
)
m <- simulate(he10_model,seed=27)
@

@@

<<m-prof-R0,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_prof_R02.rda"),seed=722,{
m_prof_R0_vals <- seq(from=25,to=35,length.out=switch(i,5,20))
tic <- Sys.time()
m_R0 <- m
m_R0_prof_u <- 1
reps <- switch(i,3,10)
param_R0_start_matrix <- matrix(coef(m_R0),byrow=TRUE,
  ncol=length(coef(m_R0)),nrow=reps*length(m_prof_R0_vals),
  dimnames=list(NULL,names(coef(m_R0))))
param_R0_start_matrix[,paste0("R0",m_R0_prof_u)] <- rep(m_prof_R0_vals,each=reps)
m_prof_R0_searches <- foreach(s=1:nrow(param_R0_start_matrix))%dopar%{
  rw_sd_call <- rep(list(0.02),times=U-1)
  names(rw_sd_call) <- lapply((1:U)[-m_R0_prof_u],function(x)paste0("R0",x))
  ibpf(m_R0,
    params=param_R0_start_matrix[s,],
    Nbpf=switch(i,2,50),Np=switch(i,10,2000),
    rw.sd=do.call(rw_sd,rw_sd_call),
    unitParNames="R0",
    sharedParNames=NULL,
    block_size=1,
    cooling.fraction.50=0.5
  )
}
toc <- Sys.time()
prof_R0_cores <- cores
})
m_prof_R0_time <- toc-tic
@

@@

<<m-slice-g,eval=T,echo=F>>=
stew(file=paste0(m_dir,"m_slice_g.rda"),seed=722,{
m_g_prof_u <- 4
slice_reps <- switch(i,2,10)
slice_points <- switch(i,5,20)
slice_g_vals <- rep(seq(from=100,to=2500,length.out=slice_points),each=slice_reps)
params_g_slice<- matrix(coef(m_g),
  nrow=slice_reps*slice_points,
  ncol=length(coef(m_g)),
  byrow=T,
  dimnames=list(NULL,names(coef(m_g))))
params_g_slice[,paste0("g",m_g_prof_u)] <- slice_g_vals
tic <- Sys.time()
m_slice_g_evals <- foreach(s=1:nrow(params_g_slice),.combine=c)%dopar%{
  logLik(bpfilter(m_g,
    params=params_g_slice[s,],
    Np=switch(i,10,5000),
    block_size=1
  ))
}
toc <- Sys.time()
m_slice_g_time <- toc-tic
m_slice_cores <- cores
})
@

@@

Figure~\ref{fig:pfilter-test-london} shows both ESS and conditional log-likelihood variance (CLLV) for measles in London.
We see that both measures agree on the most problematic observation.
However, ESS is not a sensitive proxy for CLLV elsewhere in the distribution.
CLLV can be calculated for each observation time (cumulative over all units) for a block particle filter, but for diagnostic purposes we write the conditional log-likelihood as a sum of block conditional log-likelihood (BCLL) values.
Using the notation of \citet{asfaw23arxiv}, the weights for block ${\mathcal B}_k\subset \seq{1}{U}$ are
\begin{equation}
w^j_{k,n} = \prod_{u \in {\mathcal B}_k}
            f_{Y_{n,n}|X_{u,n}}
            \big(
              y^*_{u,n}\given X^{P,j}_{u,n} \giventh \theta
            \big).
\end{equation}
The BCLL is
\begin{equation}
\hat \ell_{k,n} = \frac{1}{J}\sum_{j=1}^J w^j_{k,n},
\end{equation}
and the block approximation to the conditional likelihood of $y^*_1,\dots,y^*_{U}$ is
\begin{equation}
\hat\ell_n = \sum_{k=1}^K \hat \ell_{k,n}.
\end{equation}
For our measles spatPomp exmple, we calculate the BCLL and its variance (BCLLV) as follows:
<<bpfilter-diag-code,eval=F,echo=T,purl=F>>=
bpf_list <- foreach(rep=1:switch(i,6,40)) %dopar% {
  bpfilter(he10_model,Np=switch(i,10,2000),block_size=1)
}
bpf_ll  <- sapply(bpf_list,function(x)x@block.cond.loglik)
dim(bpf_ll) <- c(dim(bpf_list[[1]]@block.cond.loglik),length(bpf_list))
bcll <- apply(bpf_ll,c(1,2),mean)
bcllv <- apply(bpf_ll,c(1,2),var)
@
